Deep-rooted trees have more vulnerable hydraulics, yet lower drought-induced mortality in a tropical forest community

### 2020-11-02 manuscript.rmd
Changed title and summary with deep roots instead of deep water access

---
title: "Hydraulically-vulnerable trees survive on deep-water access in a tropical forest community"
author: ''
date: ''
output:
  bookdown::word_document2:
        reference_docx: manuscriptstyle.docx
        fig_caption: yes
  bookdown::pdf_document2: default
  bookdown::html_document2:
    number_sections: no
    theme: flatly
    fig_caption: yes
bibliography: references.bib
csl: new-phytologist.csl
---

@@

Section  | Word Count
------------- | -------------
Summary | 195
Main Text  | 6659
Introduction | 860
Methods | 3370
Results | 855
Discussion | 1574
Keywords | 6

@@

- Among the mechanisms that plants employ to tolerate drought, deep water access is arguably the most commonly invoked, but under-studied. Vulnerability to embolism and hydraulic safety margins can predict mortality risk at given levels of dehydration, but deep-water access may delay plant dehydration. Here, we tested the role of deep-water access in enabling survival within a diverse tropical forest community (Panama) using a novel data-model approach. 
- We inversely modeled the effective rooting depth (ERD, as the average depth of water extraction), for `r length(erd.sp)` canopy species by linking growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and leaf hydraulic vulnerability curves. We validated ERD against existing data for stable hydrogen isotopes of trees and soil. 
- Across species, deeper ERD was associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates during extreme droughts. Species with deeper ERD spent less time below critical hydraulic thresholds indicating that trees compensate for water-stress related mortality risk through deeper water access.
- The role of deep-water access in mitigating mortality of hydraulically-vulnerable trees has important implications for our predictive understanding of forest dynamics under current and future climates.

\
Keywords: hydraulic vulnerability, deep-water access, drought-induced mortality, drought tolerance, rooting depths, tropical forest, life-history strategies, safety margins

@@

\
Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from adjustment of leaf area (deciduousness), change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation to stem water storage, deep water access, increased root production, hydraulic redistribution and embolism resistance [@McDowell.2008]. Deep water access is arguably the most long-standing, yet under-studied mechanism. As plant available water varies with depth, trees in the same forest with different rooting depths, depending on species and size [@Meinzer:1999; @Chitra-Tarak.2018; @Brum.2019], differ in their experience during a drought, and thus in their growth and mortality responses [@Chitra-Tarak.2018]. A key bottleneck in community-wide testing of this mechanism has been a lack of data in both trees' rooting or water-sourcing depths and plant available soil water at those depths. A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the world’s more than 60 thousand tree species [@Fan:2017; @Beech:2017]. Further, only a small fraction of those are tropical, even though more than 90% of the world’s tree diversity resides in the tropics [@Slik:2015]. The use of stable isotopes of water as a tracer provides an indirect measure of water sourcing depths by matching the isotopic value in xylem water to vertical variations in soil pore water. However, such data are rare [@Evaristo:2016] and their interpretation is an active area of research given unresolved issues pertaining to soil heterogeneity and the correspondence of soil water extract in the laboratory to actual plant available water [@Adams.2020; @Bowers.2020; @Deurwaerder.2020]. DNA barcoding of roots [@Jones.2011] may be used to estimate species-specific rooting depths or profiles, but DNA barcode libraries for tropical forests are still under development, and the method may be cost-prohibitive for the extent of sampling required. In general, community-scale data collection for rooting or water-sourcing depths is challenging for species-rich tropical forests.

@@

To obtain the daily dynamics of soil water potential in the whole-soil column we locally parameterized and calibrated a 1-D hydrological water balance model. We then developed a novel, empirical inverse model to estimate species effective rooting depths (ERD) for large trees of `r length(erd.sp)` species (Fig. \@ref(fig:schematic)). We define ERD as the depth at which soil water potential best explained species-specific growth dynamics. The ERD model incorporated the impact of atmospheric and soil drought on growth, and was constrained with species-specific leaf vulnerability curves by using existing data for eight species, while developing trait-based proxies for the rest (based on data for a total of 21 species). We evaluated ERD against existing stable hydrogen isotope ratios ($\mathrm{\delta}$^2^H~xylem~) for a subset of six tree species as an independent observation of ERD. We evaluated whether ERD was associated with above-ground hydraulic traits sourcing the latter from a set of rare datasets for six to eight species that overlapped with our assessment of ERD. We tested the relationship of ERD with xylem vulnerability to embolism, branch hydraulic safety margins, leaf turgor loss point and maximum stem-specific hydraulic conductivity. Finally, we analysed species-specific large tree mortality dynamics over 35 years and estimated the time a species’ spent below its critical hydraulic threshold to test whether ERD explained risk of drought-induced mortality.

@@

\
  In the 50 ha plot, all stems $ge$ 1 cm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every five year through 1985-2015 for growth, mortality as well as recruitment of new stems into the ten mm dbh size class [@Condit.2017]. This inventory represents 321 woody species across the eight censuses. Aside from a two ha patch of ~100-year-old secondary forest at the north-eastern corner of the plot the remaining 48ha represent old growth forest that was never cleared for agriculture. The plot (1000 m $\times$ 500 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001]. The plot is situated in the Conrad catchment. Soils of the plot are homogeneous with red light clays, described as the Ava soil type, accounting for 72.4% of the plot [@Baillie.2017]. The topsoil field texture is silty clay loam that gradually fines to silty clay in the subsoil. Soil is mostly free draining, but restricted subsoil permeability gives rise to temporary wet season ponding. Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere [@Croat.1978; @Leigh.1982; @Gentry.1990].

@@

\
We regressed species ERD against species $\textrm{M}_{s,t}$ for each census interval $t$ with ($n$ = `r nrow(mrate.depth.mean)`) and without deciduous species ($n$ = `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))`). To assess whether deeper ERD species escaped the experience of significant soil drought, we calculated for each species the proportion of days in the simulation period (1990-2015) during which soil water potential in the soil layer matching species ERD ($\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$) were below species $\mathrm{\Psi}$~crit~. $\mathrm{\Psi}$~crit~ was defined as $\mathrm{\Psi}$~50,leaf~.

@@

\
Soil Water Potential dynamics ($\mathrm{\Psi}_{\textrm{soil},z}$) by depth (from 0.01 through 13 m) obtained from the best-fit simulations revealed that every 5-yr census interval had an extreme hydrological drought year, but census intervals varied in terms of number of extreme years, drought intensity at depth, drought seasonality and duration (Fig. \@ref(fig:swp-droughts)). Soil droughts in census intervals 2000-2005 and 2005-2010 distinctly occurred in the wet season, with one to two extreme soil drought years each, depending on the depth location (Fig. S\@ref(fig:swp-droughts)). Across the simulation period (1990-2015) $\mathrm{\Psi}_{\textrm{soil},z}$ remained above -0.5 MPa from depths 1.7 to 13 m in the simulations (Fig. S\@ref(fig:swp-all-depths)).

@@

Among evergreen species, ERD explained `r m.evg.r2.90`, `r m.evg.r2.05` and `r m.evg.r2.10`% variation in mortality rates during census intervals 1985-1990, 2000-2005 and 2005-2010, respectively, such that species with mortality rates significantly decreased with ERD ($p$ = `r m.evg.p.90`, `r m.evg.p.05` and `r m.evg.p.10`, respectively; Fig. \@ref(fig:erd-mort-census-evg)). When deciduous species were included, ERD explained lower but significant proportion of variation in mortality in those intervals ($R$^2^ = `r m.r2[1]`, `r m.r2[2]` and `r m.r2[3]`% and $p$ = `r m.p.90`, `r m.p.05` and `r m.p.10`, respectively; Fig. \@ref(fig:erd-mort-census)). Census intervals 2000-2005 and 2005-2010 were also the periods during which extreme soil droughts distinctly occurred in the wet season, effectively extending the dry season into the wet season (Fig. \@ref(fig:swp-droughts)). $\mathrm{\Psi}_{\textrm{soil},z}$ over 1985-1990 is not used for analyses, as that was the ELM-FATES spin-up period, while reliable climate forcing data was unavailable for the prior years. However, the difference between precipitation and ET shows that 1985-1990 had three extreme deficit water years, apart from the 1983 El-Niño leading to it [@Condit.2017]. Analysis of a species' time spent under significant soil drought (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) showed that species with shallower ERD spent a greater proportion of days below critical hydraulic thresholds and increasingly so over 1995-2000, 2000-2005 and 2005-2010 (Fig. \@ref(fig:erd-time); S\@ref(fig:erd-time-evg)). Species with ERD deeper than 2 m did not cross species $\mathrm{\Psi}$~crit~ for the entire 25-yr period for which soil water potential $\mathrm{\Psi}_{\textrm{soil},z}$ was modeled (1990-2015). Soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and drought periods, which is above ($\mathrm{\Psi}$~crit~) of the most sensitive tree species (-0.28 MPa).

>>>>

---
title: "Deep-rooted trees have more vulnerable hydraulics, yet lower drought-induced mortality in a tropical forest community"
author: ''
date: ''
output:
  bookdown::word_document2:
        reference_docx: manuscriptstyle.docx
        fig_caption: yes
  bookdown::pdf_document2: default
  bookdown::html_document2:
    number_sections: no
    theme: flatly
    fig_caption: yes
bibliography: references.bib
csl: new-phytologist.csl
---

@@

Section  | Word Count
------------- | -------------
Summary | 199
Main Text  | 6659
Introduction | 860
Methods | 3370
Results | 855
Discussion | 1574
Keywords | 6

@@

- Among the mechanisms that plants employ to tolerate drought, deep-rooting is arguably the most long-standing, but under-studied. Vulnerability to embolism and hydraulic safety margins can predict mortality risk at given levels of dehydration, but deep-rooting may delay plant dehydration. Here, we tested the role of deep roots in enabling survival within a diverse tropical forest community (Panama) using a novel data-model approach. 
- We inversely modeled the effective rooting depth (ERD, as the average depth of water extraction), for `r length(erd.sp)` canopy species by linking growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and leaf hydraulic vulnerability curves. We validated ERD against existing data for stable hydrogen isotopes of trees and soil. 
- Across species, deeper ERD was associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates during extreme hydrological droughts over 35 years (1981-2015). Species with deeper ERD spent less time below critical hydraulic thresholds indicating that trees compensate for water-stress related mortality risk through deep-water access.
- The role of deep roots in mitigating mortality of hydraulically-vulnerable trees has important implications for our predictive understanding of forest dynamics under current and future climates.

\
Keywords: rooting depths, deep-water access, hydraulic vulnerability, drought-induced mortality, drought tolerance, tropical forest, safety-efficiency trade-off, hydrological droughts

@@

\
Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from adjustment of leaf area (deciduousness), change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation to stem water storage, deep-rooting, increased root production, hydraulic redistribution and embolism resistance [@McDowell.2008]. Deep-rootedness is arguably the most long-standing, yet under-studied mechanism. As plant-available water varies with depth, trees in the same forest with different rooting depths, depending on species and size [@Meinzer:1999; @Chitra-Tarak.2018; @Brum.2019], differ in their experience during a drought, and thus in their growth and mortality responses [@Chitra-Tarak.2018]. A key bottleneck in community-wide testing of this mechanism has been a lack of data in both trees' rooting or water-sourcing depths and plant available soil water at those depths. A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the world’s more than 60 thousand tree species [@Fan:2017; @Beech:2017]. Further, only a small fraction of those are tropical, even though more than 90% of the world’s tree diversity resides in the tropics [@Slik:2015]. The use of stable isotopes of water as a tracer provides an indirect measure of water sourcing depths by matching the isotopic value in xylem water to vertical variations in soil pore water. However, such data are rare [@Evaristo:2016] and their interpretation is an active area of research given unresolved issues pertaining to soil heterogeneity and the correspondence of soil water extract in the laboratory to actual plant available water [@Adams.2020; @Bowers.2020; @Deurwaerder.2020]. DNA barcoding of roots [@Jones.2011] may be used to estimate species-specific rooting depths or profiles, but DNA barcode libraries for tropical forests are still under development, and the method may be cost-prohibitive for the extent of sampling required. In general, community-scale data collection for rooting or water-sourcing depths is challenging for species-rich tropical forests.

@@

To obtain the daily dynamics of soil water potential in the whole-soil column we locally parameterized and calibrated a 1-D hydrological water balance model. We then developed a novel, empirical inverse model to estimate species effective rooting depths (ERD) for large trees of `r length(erd.sp)` species (Fig. \@ref(fig:schematic)). We define ERD as the depth at which soil water potential best explained species-specific growth dynamics. The ERD model incorporated the impact of atmospheric and hydrological drought on growth, and was constrained with species-specific leaf vulnerability curves by using existing data for eight species, while developing trait-based proxies for the rest (based on data for a total of 21 species). We evaluated ERD against existing stable hydrogen isotope ratios ($\mathrm{\delta}$^2^H~xylem~) for a subset of six tree species as an independent observation of ERD. We evaluated whether ERD was associated with above-ground hydraulic traits sourcing the latter from a set of rare datasets for six to eight species that overlapped with our assessment of ERD. We tested the relationship of ERD with xylem vulnerability to embolism, branch hydraulic safety margins, leaf turgor loss point and maximum stem-specific hydraulic conductivity. Finally, we analysed species-specific large tree mortality dynamics over 35 years and estimated the time a species’ spent below its critical hydraulic threshold to test whether ERD explained risk of drought-induced mortality.

@@

\
  In the 50 ha plot, all stems $\ge$ 1 cm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every five year through 1985-2015 for growth, mortality as well as recruitment of new stems into the ten mm dbh size class [@Condit.2017]. This inventory represents 321 woody species across the eight censuses. Aside from a two ha patch of ~100-year-old secondary forest at the north-eastern corner of the plot the remaining 48ha represent old growth forest that was never cleared for agriculture. The plot (1000 m $\times$ 500 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001]. The plot is situated in the Conrad catchment. Soils of the plot are homogeneous with red light clays, described as the Ava soil type, accounting for 72.4% of the plot [@Baillie.2017]. The topsoil field texture is silty clay loam that gradually fines to silty clay in the subsoil. Soil is mostly free draining, but restricted subsoil permeability gives rise to temporary wet season ponding. Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere [@Croat.1978; @Leigh.1982; @Gentry.1990].

@@

\
We regressed species ERD against species $\textrm{M}_{s,t}$ for each census interval $t$ with ($n$ = `r nrow(mrate.depth.mean)`) and without deciduous species ($n$ = `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))`). To assess whether deeper ERD species escaped the experience of significant hydrological drought, we calculated for each species the proportion of days in the simulation period (1990-2015) during which soil water potential in the soil layer matching species ERD ($\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$) were below species $\mathrm{\Psi}$~crit~. $\mathrm{\Psi}$~crit~ was defined as $\mathrm{\Psi}$~50,leaf~.

@@

\
Soil Water Potential dynamics ($\mathrm{\Psi}_{\textrm{soil},z}$) by depth (from 0.01 through 13 m) obtained from the best-fit simulations revealed that every 5-yr census interval had an extreme hydrological drought year, but census intervals varied in terms of number of extreme years, drought intensity at depth, drought seasonality and duration (Fig. \@ref(fig:swp-droughts)). Soil droughts in census intervals 2000-2005 and 2005-2010 distinctly occurred in the wet season, with one to two extreme hydrological drought years each, depending on the depth location (Fig. S\@ref(fig:swp-droughts)). Across the simulation period (1990-2015) $\mathrm{\Psi}_{\textrm{soil},z}$ remained above -0.5 MPa from depths 1.7 to 13 m in the simulations (Fig. S\@ref(fig:swp-all-depths)).

@@

Among evergreen species, ERD explained `r m.evg.r2.90`, `r m.evg.r2.05` and `r m.evg.r2.10`% variation in mortality rates during census intervals 1985-1990, 2000-2005 and 2005-2010, respectively, such that species with mortality rates significantly decreased with ERD ($p$ = `r m.evg.p.90`, `r m.evg.p.05` and `r m.evg.p.10`, respectively; Fig. \@ref(fig:erd-mort-census-evg)). When deciduous species were included, ERD explained lower but significant proportion of variation in mortality in those intervals ($R$^2^ = `r m.r2[1]`, `r m.r2[2]` and `r m.r2[3]`% and $p$ = `r m.p.90`, `r m.p.05` and `r m.p.10`, respectively; Fig. \@ref(fig:erd-mort-census)). Census intervals 2000-2005 and 2005-2010 were also the periods during which extreme soil droughts distinctly occurred in the wet season, effectively extending the dry season into the wet season (Fig. \@ref(fig:swp-droughts)). $\mathrm{\Psi}_{\textrm{soil},z}$ over 1985-1990 is not used for analyses, as that was the ELM-FATES spin-up period, while reliable climate forcing data was unavailable for the prior years. However, the difference between precipitation and ET shows that 1985-1990 had three extreme deficit water years, apart from the 1983 El-Niño leading to it [@Condit.2017]. Analysis of a species' time spent under significant hydrological drought (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) showed that species with shallower ERD spent a greater proportion of days below critical hydraulic thresholds and increasingly so over 1995-2000, 2000-2005 and 2005-2010 (Fig. \@ref(fig:erd-time); S\@ref(fig:erd-time-evg)). Species with ERD deeper than 2 m did not cross species $\mathrm{\Psi}$~crit~ for the entire 25-yr period for which soil water potential $\mathrm{\Psi}_{\textrm{soil},z}$ was modeled (1990-2015). Soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and drought periods, which is above ($\mathrm{\Psi}$~crit~) of the most sensitive tree species (-0.28 MPa).
